\bf ``A p-value of $<$ 0.05 was considered statistically significant'': a content analysis of published statistical methods sections

### 2021-06-28 manuscript/asa_template/asa_template_almostdone.tex
revise manuscript

An ideal statistical analysis will use appropriate methods to create
insights from the data and inform the research questions. Unfortunately
many current statistical analyses are far from ideal, with many
researchers using the wrong methods, misinterpreting the results, or
failing to adequately check their assumptions \citep{2008, Leek2017}.
Some researchers take a ``mechanistic'' approach to statistics, copying
the few methods they know regardless of their appropriateness, and then
going through the motions of the analysis \citep{Stark2018}.

Many researchers lack adequate training in research methods, and
statistics is something they do with trepidation and even ignorance
\citep{Altman1994, King2019}. However, using the wrong statistical
methods can cause real harm \citep{Altman1994, Brown2018} and bad
statistical practices are being to used abet weak science
\citep{Stark2018}. Statistical mistakes are a key source of waste in
research and partly explain the current reproducibility crisis in
science \citep{Allison2016}. Even when the correct methods are used,
many researchers fail to describe them adequately, making it difficult
to reproduce the results \citep{Ernst2017, Zhou2018}. Poor statistical
methods might not be caught by reviewers, as they may not be qualified
to judge the statistics. A recent survey of editors found that only 23\%
of health and medical journals used expert statistical review for all
articles \citep{Hardwicke2020}, which was little different from a survey
from 22 years ago \citep{Goodman1998}.

There is guidance for researchers on how to write up their statistical
methods and results. The International Committee of Medical Journal
Editors recommend that researchers should: ``Describe statistical
methods with enough detail to enable a knowledgeable reader with access
to the original data to judge its appropriateness for the study and to
verify the reported results'' \citep{ICMJE2019}. More detailed guideance
is given by the SAMPL and EQUATOR guidelines
\citep{Lang2013, Altman2016} with the latter covering all apsects of the
paper. Both of these guidelines were led by Doug Altman, who spoke often
and for many years about the need for better statistical reporting. The
awareness and use of these guidelines could be improved. There were 256
Google Scholar citations to the SAMPL paper (as at 15 March 2021) which
is a good citation statistic for most papers, but is low considering the
millions of papers that use statistical analysis.

A potential contributor to poor statistical reporting is the temptation for researchers to re-use
descriptions of statistical methods, in an effort to make their papers resemble those of their peers 
and increase perceived chances of publication  \citep{Diong2018}. Two statisticians on this paper 
(AB and NW) have heard researchers admit
that they have copied-and-pasted their statistical methods sections from
other papers, regardless of whether they are appropriate. For this paper, we applied a text-based 
clustering method to analyse the content of published statistical methods sections.
Clustering results are then evaluated to estimate the extent that researchers are
using cut-and-paste or `boilerplate' statistical methods sections.
Boilerplate text is that ``which can be reused in new contexts or
applications without significant changes to the original''
\citep{Wikipedia}. Use of these methods sections indicates that little
thought has gone into the statistical analysis.

@@

We used two openly available data sources to find statistical methods
sections: research articles published in \emph{PLOS ONE} and study
protocols registered on the Australian and New Zealand Clinical Trials
Registry (ANZCTR). Data sources were chosen as examples of common
research outputs that include descriptions of statistical methods that
were used, or are planned, for analysing study outcomes.

\emph{PLOS ONE} is a open access mega-journal that publishes original
research across a wide range of scientific fields. Article submissions
are handled by an academic editor who selects peer reviewers based on
their self-nominated areas of expertise. Currently there are 324 academic editors out of 9,648 (3\%) 
with the keywords of "statistics (mathematics)" or "statistical methods" in their expertise list (web search on 25-May-2021,
 https://journals.plos.org/plosone/static/editorial-board). Submissions do not undergo
formal statistical review. Instead, reviewers are required to assess
submissions against several publication criteria, including whether:
``Experiments, statistics, and other analyses are performed to a high
technical standard and are described in sufficient detail''
\citep{PLOS}. All reviewers are asked the question: ``Has the
statistical analysis been performed appropriately and rigorously?'',
with the possible responses of ``Yes'', ``No'' and ``I don't know''.

All \emph{PLOS ONE} articles are freely accessible via the PLOS
Application Programming Interface (API). This enabled us to conduct
searches of full-text articles and analyse data on articles' text
content and general attributes such as publication date and field(s) of
research. Statistical methods sections were extracted using a two-step
approach:

\emph{Step 2}: \emph{PLOS ONE} does not prescribe standardised headings
to preface statistical methods sections. To address this, we performed
partial matching on available headings against frequently used terms in
initial search results: `Statistical analysis', `Statistical analyses',
`Statistical method', `Statistics', `Data analysis' and `Data analyses'.

Data were downloaded on 3 July 2020. For
records that did not pass Step 2, we reviewed where initial search terms
appeared in the full-text, to determine the proportion of
statistical methods sections that were missed under our chosen approach.

@@

For the statistical methods section, researchers are asked to provide a
``brief description'' of the sample size calculations, statistical
methods and planned analyses, although this section is not compulsory
\citep{ANZCTR}. Studies are reviewed by ANZCTR staff for completeness of
key information, which does not include the completeness of the
statistical methods sections.

All studies available on ANZCTR were downloaded on 1 February 2020 in
XML format. We used all the text available in the ``Statistical
methods'' section. We also collated basic information about the study
including the study type (interventional or observational), submission
date, number of funders and target sample size. These variables were
chosen as we believed they might influence the completeness of the
statistical methods section, because we expected larger studies and
those with funding to be more complete, and we also were interested in
changes over time.

@@

@@

Mathematical notation was converted from Unicode characters to plain
text. For example, the Unicode character corresponding to \(\theta\)
(\textless U+03B8\textgreater) was replaced with `theta'. Common symbols
outside of Unicode blocks including `\%' (percent) and `\textless{}'
(`less-than') were similarly converted into plain text. General
formatting was removed, including carriage returns, punctuation marks,
in-text references (e.g.~``{[}42{]}'') centred equations, and other
non-ASCII characters. Text contained inside brackets was retained to
maximise content for analysis, with brackets removed. Common stop words
including pronouns, contractions and selected prepositions were removed.
We retained selected stop words that, if excluded, may have changed the
context of statistical methods being described, for example `between'
and `against'.

We compiled an extensive list of statistical terms to standardise
reported descriptions of statistical methods. An initial list was
compiled by calculating individual word frequencies and identifying
relevant terms that appeared at least 100 times. Further terms were
sourced from index searches of three statistics textbooks
\citep{Diggle2013,Bland2015,Dobson2018}. Plurals (e.g.,
`chi-squares') unhyphenated (e.g., `chi square') and combined
(e.g.~`chisquare') terms were transformed to singular, hyphenated form
(e.g., `chi-square'). Common statistical tests were also hyphenated
(e.g., `hosmer lemeshow' to `hosmer-lemeshow'). The final list is
provided in Supplementary File 1.

@@

Text from statistical methods sections was analysed using Non-Negative
Matrix Factorization (NMF). NMF is an established approach for topic
modelling, and provides an effective solution for text-based clustering
when dealing with high-dimensional data
\citep[\citet{luong2019clustering}]{kim2014algorithms}. In this section,
we outline key details of NMF and its strengths compared with other
text-based clustering methods.

For \(N\) studies, let \(P \in R^{M \times N}\) denote a content matrix of text from statistical methods sections as \(M\) unique
terms. Text clustering algorithms for identifying common topics across
studies requires \(P\) to be represented with a vector space model. In
our case, unique terms in \(P\) are modelled using the tf-idf (term
frequency \(\times\) inverse document frequency) weighting schema, to
account for the relative importance of common and rare terms.

@@

To address these limitations, NMF deals with high-dimensional data by
mapping it to a lower-dimensional space. This mapping this achieved by
approximating \(P\) with two factor matrices: \(W \in R^{M \times g}\)
and \(H \in R ^{N \times g}\) \citep{aggarwal2012mining}, such that
\(P \approx WH^{T}\). The number of subgroups of common topic inferred
from the data is given by \(g\).

@@

Clustering results were visualised by word clouds to summarise
frequently occurring terms associated with topic membership. Whilst
these results were useful for inferring common themes across statistical
methods sections, they did not indicate evidence of boilerplate text
beyond the use of common terms. Follow up analysis therefore considered
similarities in text between sections assigned to the same topic.

We used the Jaccard similarity index to compare the statistical methods
section with the highest topic coherence score with all other sections
assigned to the same topic. Indices were calculated at the sentence
level, to account for instances of boilerplate text within larger
methods sections. We chose the Jaccard index as an easy to interpret
measure, which summarises the similarity between two sentences by the
number of words common to each sentence (intersection), divided by the
number of words that appeared in either sentence (union). Instances of
boilerplate text were defined by a Jaccard index of 0.9 or higher and a
difference in word count of plus or minus three words.

@@

@@

{\centering \includegraphics[width=0.49\linewidth]{figures/excluded_plosone} \includegraphics[width=0.49\linewidth]{figures/excluded_anzctr_missing}

\caption{\label{fig:consort-diagrams}Search results for PLOS ONE (left) and ANZCTR (right).}
\end{figure}

@@

\caption{Word clouds for ten topics for statistical methods sections published in ANZCTR}\label{fig:unnamed-chunk-9}
\end{figure}

The text clustering algorithm found topics that were purely sample size
calculations (topic 2, n = 1834), pilot studies (topic 5, n = 834),
safety/tolerability studies (topic 6, n = 524), intervention studies
(topic 8, n = 1020) and repeated measures ANOVA (topic 10 = 852).
Examples of boilerplate text are provided in Table
\ref{tab:anzctr-example-boilerplate} .

Some methods sections were only one word, including ``ANOVA'',
``t-test'', ``SPSS'' and even ``SSPS''. There were cases where the exact
same method section had been re-used in a different study. For example,
in topic 7 (n = 333), 232 sections stated `descriptive statistics' or
`descriptive statistics used' with no additional details provided. Other
instances with expanded descriptions of methods included topic 3 (n =
1277) outlining descriptive analyses only, and topic 6 (n = 524), for
hypothesis testing methods, statistical significance and software. 
In other cases, text had been slightly modified to account for changes
in primary and secondary outcomes. Examples of these text changes were
found in topics 2 and 4; identified instances
related to sample size calculations for patient recruitment to different
studies.

>>>>

An ideal statistical analysis will use appropriate methods to draw
insights from the data and inform the research questions. Unfortunately
many current statistical analyses are far from ideal, with 
researchers often using the wrong methods, misinterpreting the results, or
failing to adequately check their assumptions \citep{2008, Leek2017}.
Some researchers take a ``mechanistic'' approach to statistics, copying
the few methods they know regardless of their appropriateness, and then
going through the motions of the analysis \citep{Stark2018}. Acceptance of methodological illiteracy in favour of 
research quantitiy and perceived novelty is at odds
with principles of scientific inquiry, yet continues to pervade published scientific research \citep{VanCalster2021}. 
This paradox has been exemplified during the COVID-19 pandemic, which has led to unprecendented levels of published research of largely 
poor quality due to biases in conduct and reporting \citep{Glasziou2020,Wynants2020}.

Many researchers lack adequate training in research methods, and
statistics is something they do with trepidation and even ignorance
\citep{Altman1994, King2019}. However, using the wrong statistical
methods can cause real harm \citep{Altman1994, Brown2018} and bad
statistical practices are being to used abet weak science
\citep{Stark2018}. Statistical mistakes are a key source of research waste and are contributing to 
the current reproducibility crisis in
science \citep{Allison2016}. Even when the correct methods are used,
many researchers fail to describe them adequately, making it difficult
to reproduce the results \citep{Ernst2017, Zhou2018}. Poor statistical
methods might not be caught by reviewers, as they may not be qualified
to judge the statistics. A recent survey of editors found that only 23\%
of health and medical journals used expert statistical review for all
articles \citep{Hardwicke2020}, which was little different from a survey
from 22 years ago \citep{Goodman1998}.

There is guidance for researchers on how to write up their statistical
methods and results. The International Committee of Medical Journal
Editors recommend that researchers should: ``Describe statistical
methods with enough detail to enable a knowledgeable reader with access
to the original data to judge its appropriateness for the study and to
verify the reported results'' \citep{ICMJE2019}. More detailed guideance
is given by the SAMPL and EQUATOR guidelines
\citep{Lang2013, Altman2016} with the latter covering all apsects of the
paper tailored to different study designs. Both of these guidelines were led by Doug Altman, who spoke often
and for many years about the need for better statistical reporting. The
awareness and use of these guidelines could be improved. There were 256
Google Scholar citations to the SAMPL paper (as at 15 March 2021) which
is a good citation statistic for most papers, but is low considering the
millions of papers that use statistical analysis.

A potential contributor to poor statistical reporting is the temptation for researchers to re-use
descriptions of statistical methods, in an effort to make their papers resemble those of their peers 
and increase perceived chances of publication  \citep{Diong2018}. As descriptions become more common, rebuke by 
reviewers and journal editors can be difficult, as their past use may be argued by researchers as offering precedent for the 
conduct of analysis within their discipline \citep{Altman2002}. Two statisticians on this paper 
(AB and NW) have heard researchers admit
that they have copied-and-pasted their statistical methods sections from
other papers, regardless of whether they are appropriate. To investigate the extent of this practice, we applied a text-based 
clustering method to analyse the content of published statistical methods sections.
Clustering results are then evaluated to estimate the extent that researchers are
using cut-and-paste or `boilerplate' statistical methods sections.
Boilerplate text is that ``which can be reused in new contexts or
applications without significant changes to the original''
\citep{Wikipedia}. Use of these methods sections indicates that little
thought has gone into the statistical analysis.

@@

We used two openly available data sources to find statistical methods
sections: study
protocols registered on the Australian and New Zealand Clinical Trials
Registry (ANZCTR) and research articles published in \emph{PLOS ONE}. Data sources were chosen as examples of common
research outputs that include descriptions of statistical methods that
were planned or used for analysing study outcomes.

@@

For the statistical methods section, researchers are asked to provide a
brief description of all sample size calculations, statistical
methods and planned analyses, although this section is not compulsory
\citep{ANZCTR}. Studies are reviewed by ANZCTR staff for completeness of
key information, which does not include the completeness of the
statistical methods sections.

All studies available on ANZCTR were downloaded on 1 February 2020 in
XML format. For our analysis, we used all text available in the ``Statistical
methods'' section. We also collated basic information about the study
including the study type (interventional or observational), submission
date, number of funders and target sample size. These variables were
chosen as we believed they might influence the completeness of the
statistical methods section. For example, we hypothesised that larger studies and
those with funding to be more complete. We were also interested in
changes over time.

@@

\emph{PLOS ONE} is a open access mega-journal that publishes original
research across a wide range of scientific fields. Article submissions
are handled by an academic editor who selects peer reviewers based on
their self-nominated area(s) of expertise. Currently there are 324 academic editors out of 9,648 (3\%) 
with the keywords of "statistics (mathematics)" or "statistical methods" in their expertise list (web search on 25-May-2021,
 https://journals.plos.org/plosone/static/editorial-board). Submissions do not undergo
formal statistical review. Instead, reviewers are required to assess
submissions against several publication criteria, including whether:
``Experiments, statistics, and other analyses are performed to a high
technical standard and are described in sufficient detail''
\citep{PLOS}. All reviewers are asked the question: ``Has the
statistical analysis been performed appropriately and rigorously?'',
with the possible responses of ``Yes'', ``No'' and ``I don't know''.

All \emph{PLOS ONE} articles are freely accessible via the PLOS
Application Programming Interface (API). This enabled us to conduct
searches of full-text articles and analyse data on articles' text
content and general attributes such as publication date and field(s) of
research. Since \emph{PLOS ONE} does not prescribe a set article format, we applied a two-step
approach to identify statistical methods sections:

\emph{Step 2}: \emph{PLOS ONE} does not use standardised headings
to preface statistical methods sections. To address this, we performed
partial matching on available headings against frequently used terms in
initial search results: `Statistical analysis', `Statistical analyses',
`Statistical method', `Statistics', `Data analysis' and `Data analyses'.

Data were downloaded on 3 July 2020. For
records that did not pass Step 2, we reviewed where initial search terms
appeared in the full-text, to estimate the proportion of
statistical methods sections that were missed under our chosen approach.

@@

Mathematical notation was converted from Unicode characters to plain
text. For example, the Unicode character corresponding to \(\theta\)
(\textless U+03B8\textgreater) was replaced with `theta'. Common symbols
outside of Unicode blocks including `\%' (percent) and `\textless{}'
(`less-than') were similarly converted into plain text. General
formatting was removed, including carriage returns, punctuation marks,
in-text references (e.g.~``{[}42{]}'') centred equations, and other
non-ASCII characters. Bracketed text was retained with brackets removed to
maximise content for analysis. Common stop words
including pronouns, contractions and selected prepositions were removed.
We retained selected stop words that, if excluded, may have changed the
context of statistical methods being described, for example `between'
and `against'.

We compiled an extensive list of statistical terms to standardise
reported descriptions of statistical methods. An initial list was
compiled by calculating individual word frequencies and identifying
relevant terms that appeared at least 100 times. Further terms were
sourced from index searches of three statistics textbooks
\citep{Diggle2013,Bland2015,Dobson2018}. Plurals (e.g.,
`chi-squares') unhyphenated terms (e.g., `chi square') and combined terms
(e.g.~`chisquare') were transformed to singular, hyphenated form
(e.g., `chi-square'). Common statistical tests were also hyphenated
(e.g., `hosmer lemeshow' to `hosmer-lemeshow'). The final list is
provided in Supplementary File 1.

@@

Text from statistical methods sections was analysed using Non-Negative
Matrix Factorization (NMF). NMF is an established approach for topic
modelling, and provides an effective solution for text-based clustering
when dealing with high-dimensional data
\citep[\citet{luong2019clustering}]{kim2014algorithms}.
%In this section,
%we outline key details of NMF and its strengths compared with other
%text-based clustering methods.

For \(N\) studies, let \(P \in R^{M \times N}\) denote a content matrix of text from statistical methods sections, comprising of \(M\) unique
terms. Text clustering algorithms for identifying common topics across
studies requires \(P\) to be represented with a vector space model. In
our case, unique terms in \(P\) are modelled using the tf-idf (term
frequency \(\times\) inverse document frequency) weighting schema, to
account for the relative importance of common and rare terms.

@@

To address these limitations, NMF deals with high-dimensional data by
mapping it to a lower-dimensional space. This mapping this achieved by
approximating \(P\) with two factor matrices: \(W \in R^{M \times g}\)
and \(H \in R ^{N \times g}\) \citep{aggarwal2012mining}, such that
\(P \approx WH^{T}\). The number of subgroups of common topics inferred
from the data is given by \(g\).

@@

Results were visualised by word clouds to summarise
frequently occurring terms associated with topic membership. Follow up analysis considered 
similarities in text between statistical methods sections assign to the same topic.

The Jaccard similarity index was used to compare the statistical methods
section with the highest topic coherence score with all other sections
assigned to the same topic. Indices were calculated at the sentence
level, to account for instances of boilerplate text within larger
methods sections. We chose the Jaccard index as an easy to interpret
measure, which summarises the similarity between two sentences by the
number of words common to each sentence (intersection), divided by the
number of words that appeared in either sentence (union). Instances of
boilerplate text were defined by a Jaccard index of 0.9 or higher and a
difference in word count of plus or minus three words.

@@

{\centering \includegraphics[width=0.49\linewidth]{figures/excluded_anzctr_missing}  \includegraphics[width=0.49\linewidth]{figures/excluded_plosone}

\caption{\label{fig:consort-diagrams}Search results for ANZCTR (left) and PLOS ONE (right).}
\end{figure}

\caption{Word clouds for ten topics for statistical methods sections published in ANZCTR}\label{fig:anzctr-wordclouds}
\end{figure}

Figure \ref{fig:anzctr-wordclouds} shows word clouds for the ten topics identified by the NMF algorithm. 
Frequent words related to the largest topic (topic 2, n = 1834) reflected descriptions of sample size calculations
 and represented 19$\%$ of studies analysed. Other topics indicated pilot studies (topic 5, n = 834),
safety/tolerability studies (topic 6, n = 524), descriptive statistics (topic 7, n = 333),
intervention studies (topic 8, n = 1020) and repeated measures ANOVA (topic 10 = 852). A review of word counts by topic 
identified statistical methods sections that were only one word, including ``ANOVA'',
``t-test'', ``SPSS'' and even ``SSPS''.

There were cases where the exact
same method section had been re-used in a different study. For example,
in topic 7 (n = 333), 232 sections stated `descriptive statistics' or
`descriptive statistics used' with no additional details provided. Other
instances with expanded descriptions of methods included topic 3 (n =
1277) outlining descriptive analyses only, and topic 6 (n = 524), for
hypothesis testing methods, statistical significance and software. 
In other cases, text had been slightly modified to account for changes
in primary and secondary outcomes. Examples of these text changes were
found in topics 2 and 4; identified instances
related to sample size calculations for patient recruitment to different
studies.

Examples of boilerplate text are provided in Table
\ref{tab:anzctr-example-boilerplate}.

@@

@@
